Practical Concurrent Priority Queues \\ Student Paper for Seminar in Algorithms 2013W \\ Technical University of Vienna

### 2014-04-04 paper/paper.Rnw
Paper tweaks

\title{State of the Art in Concurrent Priority Queues \\
       Student Paper for Seminar in Algorithms 2013W \\
       Technical University of Vienna}
\author{Jakob Gruber, 0203440}

@@

% Priority queues.

Priority queues are an abstract data structure allowing insertion of items with a given priority
and removal of the lowest-priority item in logarithmic time. Search trees and Heaps (which are
flattened representations of complete trees such that each node's value is at least as large as
those of both children) are concrete data structures which are usually used to implement sequential
priority queues. However, both require fairly elaborate reorganization after \lstinline|DeleteMin|
and/or \lstinline|Insert| operations, which are especially challenging to achieve in a concurrent
environment.

@@

/* All operations are expected O(log n) time. */
void slist_insert(slist_t *l, key_t k, value_t v);
bool slist_delete(slist_t *l, key_t k, value_t &v);
\end{lstlisting}
\caption{Basic SkipList structure and operations.}
\label{fig:basicsl}
\end{figure}

@@

However, significant limitations to scalability remain. A global lock is required
to protect accesses to a variable storing the Heap's size which all operations
must obtain for a short time. Disjoint-access through bit-reversal breaks down
once a certain amount of traffic is reached, since only subsequent insertions
are guaranteed to take disjoint paths towards the root node. Note also that
the root node is a severe serial bottleneck, since it is potentially part of
every insertion path, and necessarily of every \lstinline|DeleteMin| operation.
Finally, in contrast to later dynamic SkipList-based designs, the capacity of Hunt Heap
is fixed upon creation.

@@

\citeauthor{sundell2003fast} proposed the first lock-free concurrent priority queue in
\citeyear{sundell2003fast} \cite{sundell2003fast}. The data structure is linearizable
and is implemented using commonly available atomic primitives \ac{CAS}, \ac{TAS}, and \ac{FAA}.
In contrast to other structures covered in this paper, this priority queue is restricted to
contain items with pairwise distinct priorities. Inserting a new item with a priority already contained
in the list simply performs an update of the associated value.
A real-time version is also provided which we will not discuss further (interested readers are
referred to \cite{sundell2003fast}).

@@

All remaining implementations behave similarly, displaying very low throughput under concurrency.
This is a surprise in the case of \textit{Noble}; the original benchmarks performed by
\citeauthor{sundell2003fast} led us to expect improvements upon both\textit{GlobalLock}
and \textit{Heap} \cite{sundell2003fast}. However, we are somewhat reassured in our benchmarks by
the fact that \citeauthor{linden2013skiplist}'s results are similar.

>>>>

\title{Practical Concurrent Priority Queues \\
       Student Paper for Seminar in Algorithms 2013W \\
       Technical University of Vienna}
\author{Jakob Gruber, 0203440}

@@

% Priority queues.
% TODO: Check for whitespace error @ beginning of line this paragraph.
Priority queues are an abstract data structure allowing insertion of items with a given priority
and removal of the lowest-priority item in logarithmic time. Search trees and Heaps (which are
flattened representations of complete trees such that each node's key is at least as large as
those of both children) are concrete data structures which are usually used to implement sequential
priority queues. However, both require fairly elaborate reorganization after \lstinline|DeleteMin|
and/or \lstinline|Insert| operations, which are especially challenging to achieve in a concurrent
environment.

@@

/* All operations are expected O(log n) time. */
void slist_insert(slist_t *l, key_t k, value_t v);
bool slist_delete(slist_t *l, key_t k, value_t &v);
bool slist_contains(slist_t *l, key_t k);
\end{lstlisting}
\caption{Basic SkipList structure and operations.}
\label{fig:basicsl}
\end{figure}

@@

However, significant limitations to scalability remain. A global lock is required
to protect accesses to a variable storing the Heap's size which all operations
must obtain for a short time. Disjoint-access through bit-reversal breaks down
once a certain amount of traffic is reached, since only subsequent insertions
are guaranteed to take disjoint paths towards the root node. Note also that
the root node is a severe serial bottleneck, since it is potentially part of
every insertion path, and necessarily of every \lstinline|DeleteMin| operation.
Finally, in contrast to later dynamic SkipList-based designs, the capacity of Hunt Heaps
is fixed upon creation.

@@

\citeauthor{sundell2003fast} proposed the first lock-free concurrent priority queue in
\citeyear{sundell2003fast} \cite{sundell2003fast}. The data structure is linearizable
and is implemented using commonly available atomic primitives \ac{CAS}, \ac{TAS}, and \ac{FAA}.
In contrast to other structures covered in this paper, this priority queue is restricted to
contain items with distinct priorities. Inserting a new item with a priority already contained
in the list simply performs an update of the associated value.
A real-time version is also provided which we will not discuss further (interested readers are
referred to \cite{sundell2003fast}).

@@

All remaining implementations behave similarly, displaying very low throughput under concurrency.
This is a surprise in the case of \textit{Noble}; the original benchmarks performed by
\citeauthor{sundell2003fast} led us to expect improvements upon both \textit{GlobalLock}
and \textit{Heap} \cite{sundell2003fast}. However, we are somewhat reassured in our benchmarks by
the fact that \citeauthor{linden2013skiplist}'s results are similar.
